## Introduction

Welcome to the Survival Analysis module!  This document will be periodically udated throughout the course.  As always, first thing we need to do is get all of the packages needed for the course:

```{r packages,message=F}
library(survival)
library(foreign)
library(ggplot2)
library(survminer)
library(rms)
library(flexsurv)
library(dplyr)
library(ciTools)
library(here)
library(visreg)
library(cmprsk)
```

Also need to get data sets that will be used throughout this course (all of these can be found on the GitHub repository for the course): 

```{r data sets}
loyalty=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\loyalty.csv",header=T)
recid=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\recid.csv",header=T)
recid_long=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\recid_long.csv",header=T)
recid_lag=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\recid_lag.csv",header=T)
leaders = read.csv(file = "Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\leaders.csv", header = TRUE)
bladder = read.csv(file = "Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\bladder.csv", header = TRUE)
simple=data.frame(matrix(c(7,8,10,3,2,3,1,1,0,1,1,0),ncol=2))
colnames(simple)=c("tenure","censored")

```

To perform a survival analysis, you need to identify which variable has the "time" information and which variable contains the "censoring" information.  This is done through the Surv function.  We should always visualize our data and one of the most useful visualizations is the Kaplan-Meier curve (also shown in the code below).

```{r survival curves}
# Create a Survival Analysis Object 
simple.s=Surv(time=simple$tenure,event=simple$censored)


# Create a Kaplan-Meier Survival Curve with Censoring 
simple_km=survfit(Surv(time = tenure, event = censored)~1,                      data = simple)
summary(simple_km)
plot(simple_km, main = "Survival Function", xlab = "Tenure", ylab = "Survival Probability")

loyalty.fit=survfit(Surv(Tenure, censored)~1,data=loyalty)
summary(loyalty.fit)
plot(loyalty.fit)

recid.fit = survfit(Surv(week, arrest)~1,data=recid)
summary(recid.fit)
ggsurvplot(recid.fit, data = recid, conf.int = T, palette = "purple", xlab = "Week", ylab = "Survival Probability", legend = "none", break.y.by = 0.1)


```

## Stratified Analysis

We can look at Survival curves segmented into different groups or strata (if we want to see if the likelihood of survival changes based on what group you are in).  You need to define the variable that creates the strata in order to create this analysis.  As always, to truly see if there is a difference, you should perform a statistical analysis test (there are two different tests illustrated in the following code).

```{r stratified analysis}
Loyal.KP2 = survfit(Surv(Tenure, censored) ~ Loyalty,data=loyalty)
ggsurvplot(Loyal.KP2,data=loyalty,palette = c("blue","orange"),conf.int = T)

Recid.KP = survfit(Surv(week, arrest) ~ wexp,data=recid)
ggsurvplot(Recid.KP,data=recid,palette = c("blue","orange"),conf.int = T,legend.title = "work experience", legend.labs = c("no", "yes"))

# Test for Differences in Survival Curves #
survdiff(Surv(Tenure, censored) ~ Loyalty, data=loyalty, rho=0) 
# Log-Rank Test 
survdiff(Surv(Tenure, censored) ~ Loyalty, data=loyalty, rho=1) 
# Wilcoxon Test 

survdiff(Surv(week, arrest) ~ wexp, data=recid,rho=0) # Log-Rank Test 
survdiff(Surv(week, arrest) ~ wexp,data=recid,rho=1) # Wilcoxon Test 

```

## Hazard function

We are able to calculate hazard probabilities and cumulative hazard functions in R.

```{r hazard functions}

# Calculating Hazard Probabilities 
h= simple_km$n.event/simple_km$n.risk
index.h=rep(0,length=(max(simple$tenure)+1)) #Need to add 0
index.h[(simple_km$time)+1]=h #Because of 0
haz.plot=data.frame(cbind(seq(0,max(simple$tenure)), index.h))
colnames(haz.plot)=c("Time","Hazard")
ggplot(haz.plot,aes(x=Time,y=Hazard))+geom_line()

h = loyalty.fit$n.event/loyalty.fit$n.risk
index.h=rep(0,length=(max(loyalty$Tenure)+1)) #Need to add 0
index.h[(loyalty.fit$time)+1]=h #Because of 0
haz.plot=data.frame(cbind(seq(0,max(loyalty$Tenure)), index.h))
colnames(haz.plot)=c("Time","Hazard")
ggplot(haz.plot,aes(x=Time,y=Hazard))+geom_line()

h = recid.fit$n.event/recid.fit$n.risk
index.h=rep(0,length=(max(recid$week)+1)) #Need to add 0
index.h[(recid.fit$time)+1]=h #Because of 0
haz.plot=data.frame(cbind(seq(0,max(recid$week)), index.h))
colnames(haz.plot)=c("Time","Hazard")
ggplot(haz.plot,aes(x=Time,y=Hazard))+geom_line()

ggsurvplot(recid.fit, data = recid, fun = "cumhaz", conf.int = TRUE,  palette = "purple", xlab = "Week",           ylab = "Cumulative Hazard", legend = "none")


###Cumulative hazard function

h= simple_km$n.event/simple_km$n.risk
index.h=rep(0,length=(max(simple$tenure)+1)) #Need to add 0
index.h[(simple_km$time)+1]=h #Because of 0
cum.haz=cumsum(index.h)
haz.plot=data.frame(cbind(seq(0,max(simple$tenure)), cum.haz))
colnames(haz.plot)=c("Time","Hazard")
ggplot(haz.plot,aes(x=Time,y=Hazard))+geom_line()+labs(y="Cumulative Hazard")

ggsurvplot(
  Recid.KP,
  data = recid,
  size = 1,                 
  palette =
    c("blue","orange"),
  conf.int = TRUE,          
  pval = TRUE,              
  risk.table = TRUE,        
  risk.table.col = "wexp",
  legend.labs =
    c("No", "Yes"),    
  risk.table.height = 0.25, 
  ggtheme = theme_bw()      
)


```

