
# Cox Survival Regression


```{r packages,include=F}
library(survival)
library(foreign)
library(ggplot2)
library(survminer)
library(rms)
library(flexsurv)
library(dplyr)
library(ciTools)
library(here)
library(visreg)
library(cmprsk)

library(reticulate)

use_python("C:\\ProgramData\\Anaconda3\\python.exe")


```




```{r, include=FALSE}
loyalty=read.csv("Q:/My Drive/Fall 3 - Survival Analysis/loyalty.csv",header=T)
recid=read.csv("Q:/My Drive/Fall 3 - Survival Analysis/recid.csv",header=T)
recid_long=read.csv("Q:/My Drive/Fall 3 - Survival Analysis/recid_long.csv",header=T)
recid_lag=read.csv("Q:/My Drive/Fall 3 - Survival Analysis/recid_lag.csv",header=T)
leaders = read.csv(file = "Q:/My Drive/Fall 3 - Survival Analysis/leaders.csv", header = TRUE)
bladder = read.csv(file = "Q:/My Drive/Fall 3 - Survival Analysis/bladder.csv", header = TRUE)

```


The following code fits the Proportional Cox Hazard model.

```{r Cox Regression}

# Proportional Hazards Model #
recid.ph <- coxph(Surv(week, arrest) ~ fin + age + wexp + mar +paro + prio, data = recid)
summary(recid.ph)

# Parameter Interpretation #
recid.ph2 <- coxph(Surv(week, arrest) ~ fin + age + prio, data = recid)
summary(recid.ph2)

(exp(coef(recid.ph2))-1)*100

# Automatic Selection Techniques #
full.model <- coxph(Surv(week, arrest) ~ fin + age + wexp + mar +paro + prio, 
                    data = recid)

empty.model <- coxph(Surv(week, arrest ) ~ 1, data = recid)

step.model <- step(empty.model, 
                      scope = list(lower=formula(empty.model), 
                                   upper=formula(full.model)), 
                      direction = "both")
summary(step.model)

full.model <- coxph(Surv(week, arrest) ~ fin + age  + wexp + mar +paro + prio, 
                    data = recid)

back.model <- step(full.model, direction = "backward")
summary(back.model)

full.model <- coxph(Surv(week, arrest) ~ fin + age  + wexp + mar +paro + prio, 
                    data = recid)

empty.model <- coxph(Surv(week, arrest) ~ 1, data = recid)

for.model <- step(empty.model, 
                  scope = list(lower=formula(empty.model), 
                               upper=formula(full.model)), 
                  direction = "forward")
summary(for.model)

# Estimated Survival Curves #
newdata <- data.frame(fin = c(1, 0), age = 30, wexp = c(1, 0),
                      mar = 0, paro = 0, prio = c(0, 4))

ggsurvplot(survfit(recid.ph, newdata), data = newdata, break.y.by = 0.1,
           palette = c("purple", "black"), ylab = "survival probability",
           xlab = "week", legend.labs = c("1", "2"), legend.title = "subject")


```


### Python code for Cox regression

```{python}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from lifelines import CoxPHFitter

recid = r.recid

recid_ph=recid[['fin','age', 'prio','arrest','week']]
cph = CoxPHFitter()
cph.fit(recid_ph, duration_col='week', event_col='arrest')

cph.print_summary()

cph.plot()
```


```{python}
cph.plot_partial_effects_on_outcome(covariates='prio', values=[0, 2, 4, 6, 8, 10], cmap='coolwarm')
```


## Diagnostics

Assumptions made for the Cox PH model is linearity for the continuous variables and proportional hazards throughout time (i.e. at EACH time point, the proportion between the hazards remains the same!).  We can assess both of these through use of residuals.

```{r PH diagnostics}

# Concordance 
recid.ph = coxph(Surv(week, arrest) ~ fin + age + prio, data = recid)
concordance(recid.ph)

## Martingale residuals...Linearity
## ONLY fit continuous variables here!!
recid.lin <- coxph(Surv(week, arrest) ~  age + prio, data = recid)
survminer::ggcoxfunctional(recid.lin,data=recid)

## Try transformations or can bin
recid.lin <- coxph(Surv(week, arrest) ~  sqrt(age) , data = recid)
survminer::ggcoxfunctional(recid.lin,data=recid)


# Proportional Hazard Test - Schoenfeld Residuals 
recid.ph.zph <- cox.zph(recid.ph, transform = "identity")
recid.ph.zph

ggcoxzph(recid.ph.zph)

# Can try transformations..

recid.ph.zph3 = cox.zph(recid.ph, transform = "identity")
recid.ph.zph3

recid.ph.zph4 = cox.zph(recid.ph, transform = "log")
recid.ph.zph4

```

If we believe that the PH assumption is not true for one (or more) variables, then we can explore what type of relationship is the best (with respect to time).  Options are "identity", "log" and "KM" (although identity and log are the most commonly used).  The first part of the code below illustrates how to fit the time-dependent coefficients.

The second part of the R code below illustrates how to fit  time-varying variables (variables that change their values throughout time).


```{r Time-varying information}

# Time-Dependent Coefficients #
recid.ph.tdc <- coxph(Surv(week, arrest) ~ fin +  wexp + mar + paro + prio+ age + tt(age), data = recid,
                 tt = function(x, time, ...){x*log(time)})
summary(recid.ph.tdc)

recid.ph.tdc <- coxph(Surv(week, arrest) ~  fin + prio +  age + tt(age), data = recid,
        tt = function(x, time, ...){x*log(time)})
summary(recid.ph.tdc)


```

The following code is for time-varying variables (values of the variable that change over time).  Notice that we are using a different data set (needs to be structured differently).

```{r}
# Time Varying Variables #
recid_long.ph <- coxph(Surv(start, stop, arrested) ~ fin + age + prio + employed, data = recid_long)
summary(recid_long.ph)

recid_lag.ph <- coxph(Surv(start, stop, arrested) ~ fin + age +  prio + employed, data = recid_lag)
summary(recid_lag.ph)
```


### Python for Diagnostics

For testing proportional hazard assumption:

```{python}
cph.check_assumptions(recid_ph, p_value_threshold = 0.05)
```

Python cannot program time-varying coefficients.  However, it is capable of doing time-varying variables:

```{python}
from lifelines import CoxTimeVaryingFitter

recid_1=r.recid_long
recid_long=recid_1[['fin', 'age', 'prio', 'employed','start','stop','subj','arrested']]
ctv = CoxTimeVaryingFitter()
ctv.fit(recid_long, id_col="subj", event_col="arrested", start_col="start", stop_col="stop", show_progress=True)
ctv.print_summary()
```

