
# Cox Regression


```{r packages,include=F}
library(survival)
library(foreign)
library(ggplot2)
library(survminer)
library(rms)
library(flexsurv)
library(dplyr)
library(ciTools)
library(here)
library(visreg)
library(cmprsk)
loyalty=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\loyalty.csv",header=T)
recid=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\recid.csv",header=T)
recid_long=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\recid_long.csv",header=T)
recid_lag=read.csv("Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\recid_lag.csv",header=T)
leaders = read.csv(file = "Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\leaders.csv", header = TRUE)
bladder = read.csv(file = "Q:\\My Drive\\Fall 3 2016 Survival Analysis\\Data\\Data_R\\bladder.csv", header = TRUE)
simple=data.frame(matrix(c(7,8,10,3,2,3,1,1,0,1,1,0),ncol=2))
colnames(simple)=c("tenure","censored")

```


The following code fits the Proportional Cox Hazard model.

```{r Cox Regression}

# Proportional Hazards Model #
recid.ph <- coxph(Surv(week, arrest) ~ fin + age + wexp + mar +paro + prio, data = recid)
summary(recid.ph)

# Parameter Interpretation #
recid.ph2 <- coxph(Surv(week, arrest) ~ fin + age + prio, data = recid)
summary(recid.ph2)

(exp(coef(recid.ph2))-1)*100

# Automatic Selection Techniques #
full.model <- coxph(Surv(week, arrest) ~ fin + age + wexp + mar +paro + prio, 
                    data = recid)

empty.model <- coxph(Surv(week, arrest ) ~ 1, data = recid)

step.model <- step(empty.model, 
                      scope = list(lower=formula(empty.model), 
                                   upper=formula(full.model)), 
                      direction = "both")
summary(step.model)

full.model <- coxph(Surv(week, arrest) ~ fin + age  + wexp + mar +paro + prio, 
                    data = recid)

back.model <- step(full.model, direction = "backward")
summary(back.model)

full.model <- coxph(Surv(week, arrest) ~ fin + age  + wexp + mar +paro + prio, 
                    data = recid)

empty.model <- coxph(Surv(week, arrest) ~ 1, data = recid)

for.model <- step(empty.model, 
                  scope = list(lower=formula(empty.model), 
                               upper=formula(full.model)), 
                  direction = "forward")
summary(for.model)

# Estimated Survival Curves #
newdata <- data.frame(fin = c(1, 0), age = 30, wexp = c(1, 0),
                      mar = 0, paro = 0, prio = c(0, 4))

ggsurvplot(survfit(recid.ph, newdata), data = newdata, break.y.by = 0.1,
           palette = c("purple", "black"), ylab = "survival probability",
           xlab = "week", legend.labs = c("1", "2"), legend.title = "subject")


```


## Diagnostics

Assumptions made for the Cox PH model is linearity for the continuous variables and proportional hazards throughout time (i.e. at EACH time point, the proportion between the hazards remains the same!).  We can assess both of these through use of residuals.

```{r PH diagnostics}

# Concordance 
recid.ph = coxph(Surv(week, arrest) ~ fin + age + prio, data = recid)
concordance(recid.ph)

## Partial residuals...Linearity
recid.ph <- coxph(Surv(week, arrest) ~ fin + age + wexp + mar +paro + prio, data = recid)

visreg(recid.ph, "age", xlab = "age", ylab = "partial residuals",        gg = TRUE, band = FALSE) +  geom_smooth(col = "red", 
    fill = "red") + theme_bw() 


visreg(recid.ph, "prio", xlab = "#prior convictions", ylab = "partial residuals",    gg = TRUE, band = FALSE) +  geom_smooth(col = "red", fill = "red") + theme_bw()

## Martingale residuals...Linearity
recid.lin <- coxph(Surv(week, arrest) ~  age + prio, data = recid)
survminer::ggcoxfunctional(recid.lin,data=recid)

## Try transformations
recid.lin <- coxph(Surv(week, arrest) ~  sqrt(age) , data = recid)
survminer::ggcoxfunctional(recid.lin,data=recid)


# Proportional Hazard Test - Schoenfeld Residuals 
recid.ph.zph <- cox.zph(recid.ph, transform = "identity")
recid.ph.zph

ggcoxzph(recid.ph.zph)

# Can try transformations..
recid.ph.zph2 = cox.zph(recid.ph, transform = "km")
recid.ph.zph2

recid.ph.zph3 = cox.zph(recid.ph, transform = "identity")
recid.ph.zph3

recid.ph.zph4 = cox.zph(recid.ph, transform = "log")
recid.ph.zph4

```

If we believe that the PH assumption is not true for one (or more) variables, then we can explore what type of relationship is the best (with respect to time).  Options are "identity", "log" and "KM" (although identity and log are the most commonly used).  The first part of the code below illustrates how to fit the time-dependent coefficients.

The second part of the R code below illustrates how to fit  time-varying variables (variables that change their values throughout time).


```{r Time-varying information}

# Time-Dependent Coefficients #
recid.ph.tdc <- coxph(Surv(week, arrest) ~ fin +  wexp + mar + paro + prio+ age + tt(age), data = recid,
                 tt = function(x, time, ...){x*log(time)})
summary(recid.ph.tdc)

recid.ph.tdc <- coxph(Surv(week, arrest) ~  fin + prio +  age + tt(age), data = recid,
        tt = function(x, time, ...){x*log(time)})
summary(recid.ph.tdc)

# Time Varying Variables #
recid_long.ph <- coxph(Surv(start, stop, arrested) ~ fin + age + prio + employed, data = recid_long)
summary(recid_long.ph)

recid_lag.ph <- coxph(Surv(start, stop, arrested) ~ fin + age +  prio + employed, data = recid_lag)
summary(recid_lag.ph)


```
