# Competing Risks


```{r packages,include=F}
library(survival)
library(foreign)
library(ggplot2)
library(survminer)
library(rms)
library(flexsurv)
library(dplyr)
library(ciTools)
library(here)
library(visreg)
library(cmprsk)
loyalty=read.csv("https://raw.githubusercontent.com/sjsimmo2/Survival/master/loyalty.csv",header=T)
recid=read.csv("https://raw.githubusercontent.com/sjsimmo2/Survival/master/recid.csv",header=T)
recid_long=read.csv("https://raw.githubusercontent.com/sjsimmo2/Survival/master/recid_long.csv",header=T)
recid_lag=read.csv("https://raw.githubusercontent.com/sjsimmo2/Survival/master/recid_lag.csv",header=T)
leaders = read.csv(file = "https://raw.githubusercontent.com/sjsimmo2/Survival/master/leaders.csv", header = TRUE)
bladder = read.csv(file = "https://raw.githubusercontent.com/sjsimmo2/Survival/master/bladder.csv", header = TRUE)
simple=data.frame(matrix(c(7,8,10,3,2,3,1,1,0,1,1,0),ncol=2))
colnames(simple)=c("tenure","censored")

library(reticulate)

use_python("C:\\ProgramData\\Anaconda3\\python.exe")


```

The below code illustrates how to perform a Cox regression using a cause-specific hazard for competing risks models.

```{r}
leaders$lost = factor(leaders$lost, 0:3, labels = c("In-Office", "Constitutional", "Natural", "Non-Constitutional"))
cox_nat <- coxph(Surv(years, lost == "Natural") ~ manner + start+military + age + conflict + loginc + growth + pop + land + literacy + factor(region), data = leaders)
summary(cox_nat)

```


The following code illustrates how to do a competing risk model with Fine-Gray model.

```{r Competing Risks}

# Cumulative Incidence Functions #
  cif.dat=leaders[leaders$lost!="In-Office",]  
  cif.tenure=cif.dat$years
  cif.status=cif.dat$lost

  cif.leaders=cuminc(cif.tenure,cif.status,rho=0)
  ggcompetingrisks(cif.leaders)
# Cox Regression Competing Risks #
 tenure=leaders$years
 View(leaders)
 status.leaders=leaders$lost
 x=leaders[,3:12]
 x$africa=ifelse(leaders$region==1,1,0)
 x$asia=ifelse(leaders$region==2,1,0)
 x$latin=ifelse(leaders$region==3,1,0)
 
 gray.natural=crr(tenure,status.leaders,x,failcode="Natural")
 summary(gray.natural)

 gray.const=crr(tenure,status.leaders,x,failcode="Constitutional")
 summary(gray.const)

 gray.nonconst=crr(tenure,status.leaders,x,failcode="Non-Constitutional")
 summary(gray.nonconst)

```

## Python code for competing risks models

In Python, you can easily use the cause-specific model (in either AFT or Cox PH models) by creating a new column with the event of interest being a "1" and censoring all other events (similar to what you did in Homework 2).  However, Python is NOT able run a Fine-Gray model.

# Repeated Events

```{r Repeated Events}


# Cluster model 
bladder.rse <- coxph(Surv(start, stop, event == 1) ~ rx + number + size + cluster(id), data = bladder)
summary(bladder.rse)

# Conditional Model - Assuming Variable Effects Same Across Strata - PWP model
bladder.con <- coxph(Surv(start, stop, event == 1) ~ rx + number + size + strata(enum)+cluster(id), data = bladder)
summary(bladder.con)

# Conditional Model - Assuming Variable Effects change Across Strata - PWP model
bladder.con <- coxph(Surv(start, stop, event == 1) ~ strata(enum)*rx + strata(enum)*number + strata(enum)*size + cluster(id), data = bladder)
summary(bladder.con)


# Gap Time Model - Assuming Variable Effects Same Across Strata #
# Can easily extend the gap time model to assume variable effects different same as above. #
bladder.gap <- coxph(Surv(time = (stop - start), event == 1) ~ rx + number + size + strata(enum) +cluster(id), data = bladder)
summary(bladder.gap)

```


